---
output: html_document
editor_options:
chunk_output_type: console
---

# Basic Syntax

```{r}
"Hello, World!"
```


## Arithmetic

```{r}
1 + 1

1 + 1 * 3

(1 + 1) * 3
```


::: {.infobox .caution data-latex="{caution}"}
If your code doesn't form a complete _expression_, then R will look for the rest of on the next line.

Here's an example:

```{r, eval=FALSE}
1 +
```

R says "`1 +`... what??" and if you run the code it will output something like the following:

```{r, eval=FALSE}
> 1 +
+ 
```

If you see this, press the escape/_Esc_ key.

:::


## Comments

```{r}
# comments start with `#`
# R doesn't think comments are code!
# so we can annotate our code!

# here's a (contrived) example!
-1 * -1000 # a negative number times a negative is positive
```


## Variables

```{r}
my_first_var <- "referring to data w/ names is handy!"
my_first_var
```

## Multiple Values

```{r}
c(1, 2, 3, 4, 5, 6) # `c()` is short for "combine"

1:6 # `:` lets us create sequences

my_first_vector_var <- -10:10 # we'll explain `vector`s later,
my_first_vector_var
```

## Functions

```{r}
sqrt(x = 16)
# ^^ name of function
```

```{r}
letters
toupper(x = letters)
#       ^ parameter or formal (argument)

LETTERS
tolower(x = letters)
#           ^^^^^^^ argument (always)
```

We refer to `x = letters` as a _named_ argument because we specify the parameter (`x`) to which we're passing our argument (`letters`), but we often don't specify the name of a parameter.

```{r}
tolower(letters)
```

We can't screw up too easily since `tolower()` and `toupper()` only have one paramter (`x`), but many functions can take multiple arguments.

Let's say we have a `vector` of unsorted numbers:

```{r}
unsorted_numbers <- c(3, 2, 10, 8, 1, 4, 9, 6, 5, 7)
unsorted_numbers
```

Like most languages, R has a built-in `sort()` function we can use, which works like so:

```{r}
sort(unsorted_numbers)
```

By default, `sort()` sorts in _ascending_ order, but we oftentimes will want to sort in _descending_ (or `decreasing`) order.

Rather than having a separate function called `sort_decreasing()`, we pass an argument to `sort()`'s `decreasing` parameter.

```{r}
sort(x = unsorted_numbers, decreasing = TRUE)
```

Even though `sort()` has multiple parameters, we can still skip the bames if we pass our arguments _by position_.

```{r}
sort(unsorted_numbers, TRUE)
```

Considering that `x` is `sort()`'s first parameter, and `decreaing` is `sort()`'s second parameter, we can pass our arguments (`unsorted_numbers` and `TRUE`) in the same order and R will know what we meant.

We can also mix _positional_ and _named_ arguments, and often do.

```{r}
sort(unsorted_numbers, decreasing = TRUE)
```

You're hopefully wondering "How could we know the order of `sort()`'s parameters?" which leads us to documentation.

If you want more information on a specific function, you should [check out the documentation](https://en.wikipedia.org/wiki/RTFM), which you can do with `?` or `help()`.

Here's what that looks like for `sort()`

```{r, eval=FALSE}
?sort
```

```{r, echo=FALSE}
webshot::webshot("https://stat.ethz.ch/R-manual/R-devel/library/base/html/sort.html",
                 vheight = 200)
```

There's _a ton_ of information here, but all we're interested in at the moment is the order in which we need to pass arguments to `sort()`, which we can find in the __Arguments__ section.



::: {.infobox .tip data-latex="{tip}"}
We'll cover functions in far more detail later, but sometimes it's easier to see the how the function is defined, which we can by running `sort` without `()`.

```{r}
sort
```

Only pay attention to the first line right now, which is where you'll see the following:

```
function (x, decreasing = FALSE, ...)
```

This tells us `x` is the `sort()`'s first parameter and `decreasing` is its second parameter.
:::







