---
output: html_document
editor_options:
  chunk_output_type: console
---

# Vector Geometries


Spatial vector data represent the world as a collection of points which, for two-dimensional data, are stored as $x$ and $y$ coordinates.



```{r}
suppressPackageStartupMessages(library(tidyverse))
library(sf)
```



```{r}
theme_spatial <- theme_minimal(base_family = "serif") +
  theme(axis.text = element_blank(), axis.title = element_blank(),
        axis.ticks = element_blank(), panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))

theme_set(theme_spatial)
```


Points can be joined in order to make lines, which themselves can be joined to make polygons.


```{r, echo=FALSE}
points_df1 <- data.frame(x = c(3.2, 3, 3.8, 3.5, 3.4, 3.9),
                         y = c(4, 4.6, 4.4, 3.8, 3.6, 4.5))

points_df2 <- data.frame(x = c(0, 0, 1, 2),
                         y = c(3, 4, 5, 5))

points_sf <- rbind(points_df1, points_df2) %>% 
  st_as_sf(coords = c("x", "y")) %>% 
  mutate(shape = "Points")

lines_sf <- list(points_df1, points_df2) %>% 
  map(st_as_sf, coords = c("x", "y")) %>% 
  map(summarise) %>% 
  map(st_cast, to = "LINESTRING") %>% 
  map(mutate, shape = "Lines") %>% 
  do.call(what = rbind)

poly_sf <- list(points_df1, points_df2) %>% 
  map(st_as_sf, coords = c("x", "y")) %>% 
  map(summarise) %>% 
  map(st_convex_hull) %>% 
  map(mutate, shape = "Polygons") %>% 
  do.call(what = rbind)

shapes_sf <- list(points_sf, lines_sf, poly_sf) %>% 
  do.call(what = rbind) 

shapes_gg1 <- shapes_sf %>%
  mutate(shape = factor(shape, levels = c("Points", "Lines", "Polygons"))) %>%
  arrange(desc(shape)) %>%
  ggplot() +
  geom_sf(aes(color = shape, fill = shape), size = 2, show.legend = FALSE) +
  scale_color_manual(values = c("blue", "green", "red")) +
  scale_fill_manual(values = c("blue", "green", "red")) +
  facet_wrap(~ shape) 

shapes_gg2 <- ggplot() +
  geom_sf(fill = "red", color = NA, show.legend = FALSE, data = poly_sf) +
  geom_sf(color = "green", size = 2, show.legend = FALSE, data = lines_sf) +
  geom_sf(color = "blue", size = 2, show.legend = FALSE, data = points_sf)

shapes_gg1 +
  theme(strip.text = element_text(size = 14))
```





```{r}
practice_coords <- tibble(
  lng = c(-20, -20, -10, -10, 20, 20,  10,  10), 
  lat = c(-20, 10, 20, -10, -10,  10,  20, -20),
  lab = c("A", "B", "C", "D", "E", "F", "G", "H"),
  grp = c("a", "a", "a", "a", "b", "b", "b", "b")
)

practice_coords
```

## `POINT`

`POINT` refers to the location of a single point in space. 

Here, we use `st_as_sf()` to convert a regular `data.frame` into an `sf` object. 

* Steps:
    1. take `practice_coords`
    2. convert to `sf` object with `st_as_sf()`, providing a `character` `vector` indicating the `names` of `practice_coords` in $(x, y)$ / $(longitude, latitude)$ order
    3. `mutate()` to a add a column named `shape`, which we obtain using `st_geometry_type()`.

```{r}
point_sf <- practice_coords %>%                     # Step 1.
  st_as_sf(coords = c(x = "lng", y ="lat")) %>%     # 2.
  mutate(shape = st_geometry_type(geometry))        # 3.

point_sf
```

The data in our `lng` and `lat` columns are moved to a new `geometry` column.

```{r, fig.height=3, fig.width=6}
ggplot(data = point_sf) + 
  geom_sf(aes(color = lab), size = 5, show.legend = "point") +
  labs(title = "POINT")
```

## `MULTIPOINT`

`MULTIPOINT` refers to a collection of `POINT`s. 

* Steps:
    1. take `point_sf`
    2. using `group_by()`, group the rows together based on the values in their `grp` column
    3. `summarise()` each group, which combines the points of each group into a `MULTIPOINT`
    4. `mutate()` the `shape` column to change it to the new `st_geometry_type()`

```{r}
multi_point_sf <- point_sf %>%               # Step 1.
  group_by(grp) %>%                          # 2.
  summarise() %>%                            # 3.
  mutate(shape = st_geometry_type(geometry)) # 4.

multi_point_sf
```

Instead of the 8 separate `POINT`s with which we started, we now have 2 rows of `MULTIPOINT`s, each of which contain 4 points.

```{r, fig.height=3, fig.width=6}
ggplot(data = multi_point_sf) + 
  geom_sf(aes(color = grp), size = 5, show.legend = "point") +
  labs(title = "MULTIPOINT")
```
    
## `LINESTRING`

`LINESTRING` is how we represent individual lines.

* Steps:
    1. take `multi_point_sf`
    2. cast the `geometry` `to=` `LINESTRING` using `st_cast()`
    3 `mutate()` the `shape` column to change it to the new `st_geometry_type()`

```{r}
linestring_sf <- multi_point_sf %>%          # Step 1.
  st_cast(to = "LINESTRING") %>%             # 2.
  mutate(shape = st_geometry_type(geometry)) # 3.

linestring_sf
```

Now we have 2 rows that each contain a `LINESTRING`, which was built by connecting each point to the next.

```{r, fig.height=3, fig.width=6}
ggplot(data = linestring_sf) + 
  geom_sf(aes(color = grp), size = 2, show.legend = "line") +
  labs(title = "LINESTRING")
```

## `MULTILINESTRING`

Similar to `MULTIPOINT`s that contain multiple points, we also have `MULTILINESTRING`s.

* Steps:
    1. take `linestring_sf`
    2. `summarise()` the rows, combining them all into a single `MULTILINESTRING`
    3. `mutate()` the `shape` column to change it to the new `st_geometry_type()` and replace the `grp` column that is dropped when we `summarise()` without using `group_by()`

```{r}
multi_linestring_sf <- linestring_sf %>%     # Step 1.
  summarise() %>%                            # 2.
  mutate(shape = st_geometry_type(geometry), # 3.
         grp = "multi")                      # 4.

multi_linestring_sf
```

Now we have 2 lines embedded inside a single `MULTILINESTRING` row.

```{r, fig.height=3, fig.width=6}
ggplot(data = multi_linestring_sf) + 
  geom_sf(aes(color = grp), size = 2, show.legend = "line") +
  labs(title = "MULTILINESTRING")
```


## `POLYGON`

`POLYGON`s are essentially sets of lines that close to form a ring, although `POLGYON`s can also contain holes. We can easily wrap a shape around any `geometry` using `st_convex_hull()` to form a [convex hull](https://en.wikipedia.org/wiki/Convex_hull) polygon.

* Steps:
    1. take `point_sf`
    2. using `group_by()`, group the rows together based on the values in their `grp` column
    3. `summarise()` each group, combining them into `MULTIPOINT`s
    4. wrap the `MULTIPOINT`s in a polygon using `st_convex_hull()`
    5. `mutate()` the `shape` column to change it to the new `st_geometry_type()`

```{r}
polygon_sf <- point_sf %>%                   # Step 1.
  group_by(grp) %>%                          # 2.
  summarise() %>%                            # 3.
  st_convex_hull() %>%                       # 4.
  mutate(shape = st_geometry_type(geometry)) # 5.

polygon_sf
```


```{r, fig.height=3, fig.width=6}
ggplot(data = polygon_sf) + 
  geom_sf(aes(fill = grp), show.legend = "polygon") +
  labs(title = "POLYGON")
```


## `MULTIPOLYGON`

`POLYGON`s can also be grouped together to form `MULTIPOLYGON`s.

* Steps:
    1. take `polygon_sf`
    2. `summarise()` the rows, combining them all into a single `MULTILIPOLYGON` 
    3. `mutate()` the `shape` column to change it to the new `st_geometry_type()` and replace the `grp` column that is dropped when we `summarise()` without using `group_by()`

```{r}
multi_polygon_sf <- polygon_sf %>%           # Step 1.
  summarise() %>%                            # 2.
  mutate(shape = st_geometry_type(geometry), # 3.
         grp = "multi")

multi_polygon_sf
```

```{r, fig.height=3, fig.width=6}
ggplot(data = multi_polygon_sf) +
  geom_sf(aes(fill = grp), show.legend = "polygon") +
  labs(title = "MULTIPOLYGON")
```


## `GEOMETRY`

`GEOMETRY` is a special `geometry` type. It refers to a column of mixed geometries, i.e. we have multiple geometry types in our `geometry` column. 


```{r}
geometry_sf <- list(point_sf, multi_point_sf, linestring_sf, 
                    multi_linestring_sf, polygon_sf, multi_polygon_sf) %>% 
  map_if(~ "lab" %in% names(.x), select, -lab) %>%
  do.call(what = rbind) %>% 
  mutate(grp = if_else(shape == "POINT", as.character(row_number()), grp))

geometry_sf
```


```{r}
ggplot(data = geometry_sf) +
  geom_sf(aes(color = grp, fill = grp), size = 2, show.legend = FALSE) +
  facet_wrap(~ shape, nrow = 2) +
  labs(title = "GEOMETRY")
```




