---
output: html_document
editor_options:
  chunk_output_type: console
---

# Spatial Data

```{r}
packages <- c(
  "leaflet",  # interactive web mapping
  "osmdata",  # Open Street Maps API data
  "raster",   # obtaining administrative boundary data and spatial raster data handling
  "sf",       # spatial vector data handling
  "stars",    # {sf}'s spatio-temporal raster counterpart
  "tidyverse" # data manipulation
)

install.packages(
  packages[!sapply(packages, requireNamespace, quietly = TRUE)]
)

library(leaflet)
library(sf)
library(stars)
library(tidyverse)
```


```{r, echo=FALSE}
theme_knapp <- theme_minimal(base_family = "serif") +
  theme(axis.text = element_blank(), axis.title = element_blank(),
        axis.ticks = element_blank(), panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5))

theme_set(theme_knapp)
```

```{r, echo=FALSE}
osm_all_file_path <- "data/osm_all.rds"

if (!file.exists(osm_all_file_path)) {
  bbox <- c(70.7, 34.27, 70.9, 34.327)
  
  get_osm_data <- function(.bbox, ...) {
    query <- osmdata::add_osm_feature(osmdata::opq(.bbox), ...)
    osmdata::osmdata_sf(query)
  }
  
  osm_water <- get_osm_data(bbox, key = "natural", value = "water")
  osm_wtrwy <- get_osm_data(bbox, key = "waterway")
  osm_roads <- get_osm_data(bbox, key = "highway")
  osm_bldgs <- get_osm_data(bbox, key = "building")
  
  all_water <- list(osm_water$osm_polygons,
                    osm_wtrwy$osm_lines, osm_wtrwy$osm_multilines,
                    osm_wtrwy$osm_polygons, osm_wtrwy$osm_multipolygons) %>%
    map(select, geometry) %>%
    do.call(what = rbind)
  
  all_buildings <- osm_bldgs %>%
    map(keep, inherits, "sf") %>%
    compact() %>%
    do.call(what = rbind)
  
  colors_and_levels <- c("#0077be", "gray", "green") %>%
    set_names(c("Water", "Roads", "Prominent Buildings"))
  
  osm_all <- list(all_water, osm_roads$osm_lines, all_buildings) %>%
    map(st_crop, osm_roads$osm_lines) %>%
    map2(names(colors_and_levels),
         ~ mutate(.x, type = factor(.y, levels = names(colors_and_levels)))) %>%
    map(select, type, geometry) %>%
    do.call(what = rbind) %>%
    mutate(col = recode(type, !!!colors_and_levels))
  
  saveRDS(osm_all, osm_all_file_path)
}


osm_all <- readRDS(osm_all_file_path)

get_centroid_lab <- function(x) {
  if (!requireNamespace("dplyr", quietly = TRUE)) stop("{dplyr} is required.")
  if (!requireNamespace("sf", quietly = TRUE)) stop("{sf} is required.")
  xy_mat <- round(sf::st_coordinates(sf::st_centroid(dplyr::summarise(x))), 2)
  sprintf("EPSG: %s\nCentroid (X, Y): %s, %s",
          sf::st_crs(x)$epsg, xy_mat[[1]], xy_mat[[2]])
}

all_ggs <- osm_all %>%
  ggplot() +
  geom_sf(aes(color = col, fill = col), 
          show.legend = FALSE) +
  scale_fill_identity() +
  scale_color_identity()


all_ggs +
  facet_wrap(~ type, nrow = 1, strip.position = "top")

all_ggs +
  theme(plot.caption = element_text(hjust = 0.9, vjust = 1))
```



```{r, echo=FALSE}
elev_rast_file_path <- "data/elev_rast.rds"

if (!file.exists(elev_rast_file_path)) {
  locs <- osm_all %>% 
      summarise() %>% 
      st_convex_hull() %>% 
      st_buffer(0.2) %>% 
      as_Spatial()
  
  elev_rast <- elevatr::get_elev_raster(
    locations = locs,
    z = 9,
    clip = "bbox"
  ) %>% 
    st_as_stars() %>% 
    st_crop(st_bbox(osm_all))
  
  saveRDS(elev_rast, elev_rast_file_path)
}

elev_rast <- readRDS(elev_rast_file_path)

elev_rast$layer[102:117, 70:86] %>%
  apply(2, rev) %>%
  `colnames<-`(seq_len(ncol(.))) %>% 
  as_tibble() %>%
  mutate(row = row_number()) %>% 
  gather(var, val, -row) %>% 
  mutate(col = str_remove(var, "^V") %>% as.integer()) %>% 
  arrange(desc(col)) %>%
  ggplot(aes(x = col, y = row)) +
  geom_tile(aes(fill = val), color = "black") +
  geom_text(aes(label = round(val)), fontface = "bold", family = "mono") +
  scale_fill_gradient(low = "lightblue", high = "red") +
  guides(fill = guide_colorbar(title = NULL, barheight = 0.75, barwidth = 25)) +
  theme(legend.position = "top")
```


```{r, echo=FALSE, warning=FALSE}
ggplot() +
  geom_stars(data = elev_rast) +
  scale_fill_viridis_c(option = "inferno", na.value = NA) +
  guides(fill = guide_colorbar(title = "Elevation (m)", title.position = "top",
                               title.hjust = 0.5,
                               barheight = 0.5, barwidth = 25)) +
  theme(legend.position = "top")
```


```{r, echo=FALSE, warning=FALSE}
ggplot() +
  geom_stars(data = elev_rast, show.legend = FALSE) +
  geom_sf(aes(color = col),
          fill = "#0077be", show.legend = FALSE,
          data = osm_all) +
  scale_color_identity() +
  scale_fill_viridis_c(option = "inferno", na.value = NA)
```


## `geometry`

Spatial vector data represent the world as a collection of points which, for two-dimensional data, are stored as $x$ and $y$ coordinates. 

Points can be joined in order to make lines, which themselves can be joined to make polygons.


```{r, echo=FALSE}
points_df1 <- data.frame(x = c(3.2, 3, 3.8, 3.5, 3.4, 3.9),
                         y = c(4, 4.6, 4.4, 3.8, 3.6, 4.5))

points_df2 <- data.frame(x = c(0, 0, 1, 2),
                         y = c(3, 4, 5, 5))

points_sf <- rbind(points_df1, points_df2) %>% 
  st_as_sf(coords = c("x", "y")) %>% 
  mutate(shape = "Points")

lines_sf <- list(points_df1, points_df2) %>% 
  map(st_as_sf, coords = c("x", "y")) %>% 
  map(summarise) %>% 
  map(st_cast, to = "LINESTRING") %>% 
  map(mutate, shape = "Lines") %>% 
  do.call(what = rbind)

poly_sf <- list(points_df1, points_df2) %>% 
  map(st_as_sf, coords = c("x", "y")) %>% 
  map(summarise) %>% 
  map(st_convex_hull) %>% 
  map(mutate, shape = "Polygons") %>% 
  do.call(what = rbind)

shapes_sf <- list(points_sf, lines_sf, poly_sf) %>% 
  do.call(what = rbind) 

shapes_gg1 <- shapes_sf %>%
  mutate(shape = factor(shape, levels = c("Points", "Lines", "Polygons"))) %>%
  arrange(desc(shape)) %>%
  ggplot() +
  geom_sf(aes(color = shape, fill = shape), size = 2, show.legend = FALSE) +
  scale_color_manual(values = c("blue", "green", "red")) +
  scale_fill_manual(values = c("blue", "green", "red")) +
  facet_wrap(~ shape) 

shapes_gg2 <- ggplot() +
  geom_sf(fill = "red", color = NA, show.legend = FALSE, data = poly_sf) +
  geom_sf(color = "green", size = 2, show.legend = FALSE, data = lines_sf) +
  geom_sf(color = "blue", size = 2, show.legend = FALSE, data = points_sf)

shapes_gg1 +
  theme(strip.text = element_text(size = 14))
```





```{r}
practice_coords <- tibble(lng = c(-20, -20, -10, -10, 20, 20,  10,  10), 
                          lat = c(-20, 10, 20, -10, -10,  10,  20, -20),
                          lab = c("A", "B", "C", "D", "E", "F", "G", "H"),
                          grp = c("a", "a", "a", "a", "b", "b", "b", "b"))
practice_coords
```

#### `POINT`

`POINT` refers to the location of a single point in space. 

Here, we use `st_as_sf()` to convert a regular `data.frame` into an `sf` object. 

* Steps:
    1. take `practice_coords`
    2. convert to `sf` object with `st_as_sf()`, providing a `character` `vector` indicating the `names` of `practice_coords` in $(x, y)$ / $(longitude, latitude)$ order
    3. `mutate()` to a add a column named `shape`, which we obtain using `st_geometry_type()`.

```{r}
point_sf <- practice_coords %>%              # Step 1.
  st_as_sf(coords = c("lng", "lat")) %>%     # 2.
  mutate(shape = st_geometry_type(geometry)) # 3.
point_sf
```

The data in our `lng` and `lat` columns are moved to a new `geometry` column.

```{r, fig.height=3, fig.width=6}
ggplot(data = point_sf) + 
  geom_sf(aes(color = lab), size = 5, show.legend = "point") +
  labs(title = "POINT")
```

#### `MULTIPOINT`

`MULTIPOINT` refers to a collection of `POINT`s. 

* Steps:
    1. take `point_sf`
    2. using `group_by()`, group the rows together based on the values in their `grp` column
    3. `summarise()` each group, which combines the points of each group into a `MULTIPOINT`
    4. `mutate()` the `shape` column to change it to the new `st_geometry_type()`

```{r}
multi_point_sf <- point_sf %>%               # Step 1.
  group_by(grp) %>%                          # 2.
  summarise() %>%                            # 3.
  mutate(shape = st_geometry_type(geometry)) # 4.

multi_point_sf
```

Instead of the 8 separate `POINT`s with which we started, we now have 2 rows of `MULTIPOINT`s, each of which contain 4 points.

```{r, fig.height=3, fig.width=6}
ggplot(data = multi_point_sf) + 
  geom_sf(aes(color = grp), size = 5, show.legend = "point") +
  labs(title = "MULTIPOINT")
```
    
#### `LINESTRING`

`LINESTRING` is how we represent individual lines.

* Steps:
    1. take `multi_point_sf`
    2. cast the `geometry` `to=` `LINESTRING` using `st_cast()`
    3 `mutate()` the `shape` column to change it to the new `st_geometry_type()`

```{r}
linestring_sf <- multi_point_sf %>%          # Step 1.
  st_cast(to = "LINESTRING") %>%             # 2.
  mutate(shape = st_geometry_type(geometry)) # 3.

linestring_sf
```

Now we have 2 rows that each contain a `LINESTRING`, which was built by connecting each point to the next.

```{r, fig.height=3, fig.width=6}
ggplot(data = linestring_sf) + 
  geom_sf(aes(color = grp), size = 2, show.legend = "line") +
  labs(title = "LINESTRING")
```

#### `MULTILINESTRING`

Similar to `MULTIPOINT`s that contain multiple points, we also have `MULTILINESTRING`s.

* Steps:
    1. take `linestring_sf`
    2. `summarise()` the rows, combining them all into a single `MULTILINESTRING`
    3. `mutate()` the `shape` column to change it to the new `st_geometry_type()` and replace the `grp` column that is dropped when we `summarise()` without using `group_by()`

```{r}
multi_linestring_sf <- linestring_sf %>%     # Step 1.
  summarise() %>%                            # 2.
  mutate(shape = st_geometry_type(geometry), # 3.
         grp = "multi")                      # 4.

multi_linestring_sf
```

Now we have 2 lines embedded inside a single `MULTILINESTRING` row.

```{r, fig.height=3, fig.width=6}
ggplot(data = multi_linestring_sf) + 
  geom_sf(aes(color = grp), size = 2, show.legend = "line") +
  labs(title = "MULTILINESTRING")
```


#### `POLYGON`

`POLYGON`s are essentially sets of lines that close to form a ring, although `POLGYON`s can also contain holes. We can easily wrap a shape around any `geometry` using `st_convex_hull()` to form a [convex hull](https://en.wikipedia.org/wiki/Convex_hull) polygon.

* Steps:
    1. take `point_sf`
    2. using `group_by()`, group the rows together based on the values in their `grp` column
    3. `summarise()` each group, combining them into `MULTIPOINT`s
    4. wrap the `MULTIPOINT`s in a polygon using `st_convex_hull()`
    5. `mutate()` the `shape` column to change it to the new `st_geometry_type()`

```{r}
polygon_sf <- point_sf %>%                   # Step 1.
  group_by(grp) %>%                          # 2.
  summarise() %>%                            # 3.
  st_convex_hull() %>%                       # 4.
  mutate(shape = st_geometry_type(geometry)) # 5.

polygon_sf
```


```{r, fig.height=3, fig.width=6}
ggplot(data = polygon_sf) + 
  geom_sf(aes(fill = grp), show.legend = "polygon") +
  labs(title = "POLYGON")
```


#### `MULTIPOLYGON`

`POLYGON`s can also be grouped together to form `MULTIPOLYGON`s.

* Steps:
    1. take `polygon_sf`
    2. `summarise()` the rows, combining them all into a single `MULTILIPOLYGON` 
    3. `mutate()` the `shape` column to change it to the new `st_geometry_type()` and replace the `grp` column that is dropped when we `summarise()` without using `group_by()`

```{r}
multi_polygon_sf <- polygon_sf %>%           # Step 1.
  summarise() %>%                            # 2.
  mutate(shape = st_geometry_type(geometry), # 3.
         grp = "multi")

multi_polygon_sf
```

```{r, fig.height=3, fig.width=6}
ggplot(data = multi_polygon_sf) +
  geom_sf(aes(fill = grp), show.legend = "polygon") +
  labs(title = "MULTIPOLYGON")
```


#### `GEOMETRY`

`GEOMETRY` is a special `geometry` type. It refers to a column of mixed geometries, i.e. we have multiple geometry types in our `geometry` column. 


```{r}
geometry_sf <- list(point_sf, multi_point_sf, linestring_sf, 
                    multi_linestring_sf, polygon_sf, multi_polygon_sf) %>% 
  map_if(~ "lab" %in% names(.x), select, -lab) %>%
  do.call(what = rbind) %>% 
  mutate(grp = if_else(shape == "POINT", as.character(row_number()), grp))

geometry_sf
```


```{r}
ggplot(data = geometry_sf) +
  geom_sf(aes(color = grp, fill = grp), size = 2, show.legend = FALSE) +
  facet_wrap(~ shape, nrow = 2) +
  labs(title = "GEOMETRY")
```




